---
title: "Udacity_EDA/White_Wine"
author: "Helen Nguyen"
date: "6/7/2018"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

Wine Exploratory Data Analysis by Helen Nguyen
========================================================

This Udacity project applies EDA using R to analyze a white wine dataset. The main objective is to understand and determine which variables affect wine quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Load libraries
library(ggplot2)
library(gridExtra)
library(dplyr)
library(corrplot)
library(RColorBrewer)
library(knitr)
library(GGally)
```

```{r, Load_the_Data}
#Load the Data
ww <- read.csv('/Users/Helen/Desktop/udacityR/wineQualityWhites.csv')
```

```{r}
#Remove unnecessary X column
ww <- ww %>% dplyr::select(-X)
```

```{r echo=FALSE, Univariate_Plots1}
#Convert Quality from an Integer to a Factor 
ww$quality <- factor(ww$quality, ordered = T)
```

```{r echo=FALSE, Univariate_Plots2}
# Create a factored variable and name it 'Rating'
ww$rating <- ifelse(ww$quality < 5, 'bad', ifelse(ww$quality < 7, 'average', 'excellent'))
ww$rating <- ordered(ww$rating, levels = c('bad', 'average', 'excellent'))
```

# Univariate Plots Section

Let's take  a look at dimensions, structure, and summary of the dataset

```{r echo=FALSE, Univariate_Plots3}
#Print dimensions
dim(ww)
```

```{r echo=FALSE, Univariate_Plots4}
#Print structure
str(ww)
```

```{r echo=FALSE, Univariate_Plots5}
#Print summary
summary(ww)
```
Wine quality has a median value of 6 with a minimum of 3 and a maximum of 9.

Fixed acidity has a high maximum 14.2 while it's mean is 8.9 and minimum is 3.8.

Under residual.sugar, there is an unusually high maximum of 65.8 which means that a white wine in the dataset contains a much larger concentration of sugar than the rest. A wine with more than 45 grams/liter of sugar is considered sweet.

Density values fall between .99 and 1.

Alcohol has a mean of 10.5% with a minimum of 8.0% and a maximum of 14.2%.

Now that we have a summary of these variables, creating plots will allow us to view their distribution.

```{r echo=FALSE, Univariate_Plots6}
ggplot(aes(x=quality), data=ww) + 
  geom_bar() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Count', 
       title = 'Distribution of Quality')
```

6 is the most common value for wine quality followed by 5 and 7. This tells us that most wines are average. Only a few wines were rated at the opposite ends of the quality scale.

```{r  echo=FALSE, Univariate_Plots7}
ggplot(aes(x=rating), data=ww) + 
  geom_bar() + 
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Rating', y = 'Count', 
       title = 'Distribution of Rating')
```

A large majority of wines are rated average. Due to this finidng, it may affect the accuracy of our model.

```{r echo=FALSE, Univariate_Plots8}
ggplot(aes(x = fixed.acidity), data=ww) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(breaks = seq(4, 15, 2)) +
  geom_vline(xintercept = mean(ww$fixed.acidity, color = 'blue')) +
  geom_vline(xintercept = median(ww$fixed.acidity, color = 'yellow', linetype = 2)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Fixed Acidity (g/dm^3)', y = 'Count', 
      title = 'Distribution of Fixed Acidity')
```

Fixed acidity shows a normal distrubution with mean (6.86) and median(6.80) values that are very close.

```{r echo=FALSE, Univariate_Plots9}
#Original volatile acidity plot
p1 <- ggplot(aes(x=volatile.acidity), data = ww) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(0,1.1,0.1),limits = c(0.08,0.8)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Volatile Acidity (g/dm^3)', y = 'Count',
      title = 'Distribution of Original Volatile Acidity')

#Volatile acidity plot with log10
p2 <- ggplot(aes(x=volatile.acidity), data = ww) +
  geom_histogram(binwidth = 0.01)+
  scale_x_log10() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Volatile Acidity (g/dm^3)', y = 'Count',
      title = 'Distribution of log10(Volatile Acidity)')

grid.arrange(p1, p2, ncol = 1)
```

Volatile acidity exhibits a normal distribution after applying the log function and removing outliers.

```{r echo=FALSE, Univariate_Plots10}
ggplot(aes(x = citric.acid), data = ww) + 
  geom_histogram(binwidth = 0.02) +
  scale_x_continuous(limits = c(0, 1.0), breaks = seq(0, 1.0, .1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Citric Acid (g/dm^3)', y = 'Count',
      title = 'Distribution of Citric Acid')
```

Citric acid appears to also have a normal distribution with many outliers. 

```{r echo=FALSE, Univariate_Plots11}
#After checking range, use coor_cartersion to set the plot range
p1 <- ggplot(aes(x=residual.sugar), data = ww) +
  coord_cartesian(xlim = c(0.6, 20)) +
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  labs(x = 'Residual sugar (g/dm^3)', y = 'Count', 
  title = 'Distribution of Original Residual Sugar')

p2 <-  ggplot(aes(x=residual.sugar), data = ww) +
  coord_cartesian(xlim = c(0.6, 20)) +
  geom_histogram(binwidth = 0.05) +
  scale_x_log10(breaks = seq(0, 20, 2)) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  labs(x = 'Residual Sugar (g/dm^3)', y = 'Count', 
  title = 'Distribution of log10(Residual Sugar)')

grid.arrange(p1, p2, ncol = 1)
```

After removing outliers, residual sugar has a bimodal distribution with peaks before and after 3.

```{r echo=FALSE, Univariate_Plots12}
ggplot(aes(x = chlorides), data=ww) +
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(lim = c(0,0.1), breaks = seq(0, 0.1, .01)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Chlorides (g/dm^3)', y = 'Count',
       title = 'Distribution of Chlorides')
```

Chlorides have a normal distribution after removing outliers. Most chlorides fall between .035-.05.

```{r echo=FALSE, Univariate_Plots13}
ggplot(aes(x = free.sulfur.dioxide), data=ww) +
  geom_histogram(binwidth = 5) +
  scale_x_continuous(breaks = seq(0,90,3), limits = c(0,90)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Free Sulfur Dioxide(mg/dm^3)', y = 'Count',
       title = 'Distribution of Free Sulfur Dioxide')
```

Free sulfur dioxide exhibits a normal distribution with a median of of 34.0 and mean of 35.3.

```{r echo=FALSE, Univariate_Plots14}
ggplot(aes(x = total.sulfur.dioxide ), data=ww) +
  geom_histogram(binwidth = 5) +
  scale_x_continuous(breaks = seq(0, 300, 15), limits = c(0,300)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Total Sulfur Dioxide (mg/dm^3)', y = 'Count',
       title = 'Distribution of Total Sulfur Dioxide')
```

Total sulfur dioxode has a normal distribution with a median of 134.0 and mean of 138.4.

```{r echo=FALSE, Univariate_Plots15}
ggplot(aes(x = density), data = ww) + 
  geom_histogram(binwidth = 0.001) + 
  xlim(0.98, 1.04) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Density (g/cm^3)', y = 'Count',
       title = 'Distribution of Density')
```

Density appears to have a normal distribution and very few outliers.

```{r echo=FALSE, Univariate_Plots16}
ggplot(aes(x = pH), data=ww) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(2.7, 4, 0.1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'pH', y = 'Count',
       title = 'Distribution of pH')
```

pH has a normal distribution with a minimum of 2.7 and maximum of 3.8.

```{r echo=FALSE, Univariate_Plots17}
ggplot(aes(x = sulphates), data=ww) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(lim = c(0.2,1.1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Sulphates (g/dm^3)', y = 'Count',
       title = 'Distribution of Sulphates')
```

Sulphates exhibit a long tailed distribution with very close median (.47) and mean (.49) values. 

```{r echo=FALSE, Univariate_Plots18}
ggplot(aes(x = alcohol), data=ww) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(lim = c(8,14)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Alcohol (% by volume)', y = 'Count',
       title = 'Distribution of Alcohol')
```

Alcohol does not have a normal distribution with multiple peaks between 8.0% and 14.2%.

# Univariate Analysis

### Dataset Structure 

The dataset contains 4,898 wines along with 11 quantitative variables. In addition to these variables, the dataset also includes quality and rating (created above). At least 3 wine experts rated the quality of each wine with a rating between 0 (very bad) and 10 (very excellent).

Most wines appear to be average quality with a few being very bad or excellent. Due to this finding, it may be difficult to create a predictive model since there's not enough data on excellent and bad wines.

### Main Feature in Dataset

The main feature I'm interested in is quality. The goal is to determine which variables affect wine quality.

### Hypothesis 

I think alcohol and residual sugar have an impact on wine quality.

### New Variable

I created an additional variable called "rating" to label wines as either bad, average, or excellent.

### Unusual Distributions

Volatile acidity and citric acid have a significant number of outliers. I had to apply a log function to remove the outliers for volatile acidity to see that it has a normal distribution. 

# Bivariate Plots Section

Now that we've looked at these variables individually, let's take a look at the correlation between them.

```{r echo=FALSE, Bivariate_Plots1}
#Create new variable to store quality as a number
ww$numQuality = ww$quality
```

```{r echo=FALSE, Bivariate_Plots2}
#Convert wine quality to a number
ww$quality = as.numeric(ww$quality)
```

```{r Correlations, echo=FALSE, fig.width=20, fig.height=20}
ggcorr(ww, 
       method = c('all.obs', 'spearman'),
       nbreaks = 4, palette = 'Pu0r', label = TRUE,
       name = 'spearman correlation coeff.(rho)',
       hjust =0.8, angle =-70, size=3) +
  ggtitle('Spearman Correlation Coefficient Matrix')
```

Correlation with Quality

• Alcohol is the most positively correlated with quality (.4)
• Density and quality have the strongest negative correlation (-.3)
• Residual sugar and quality have a sliglyly negative correlation (-.1)
• Volatile acidity and fixed acidity have a slightly negative correlation (-.2) with quality
• Sulphates and quality have a slightly postive correlation.

Correlation among Variables

• Density and residual sugar have the strongest positive correlation (.8)
• Density and alcohol have the strongest negative correlation  (-.8)
• Alcohol and residual sugar have a negative correlation (-.5)

In addition to the correlation table, boxplots will help us explore the relationships among these variables.

```{r echo=FALSE, Bivariate_Plots4}
#Convert quality to a factor 
ww$quality = as.factor(ww$quality)
```

```{r echo=FALSE, Bivariate_Plots5}
ggplot(aes(x = quality, y = alcohol), data = ww) + 
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Alcohol (% by volume)',
       title = 'Relationship between Alcohol and Quality')
```

As we saw in the correlation table, alcohol and quality have a strong positive relationship. This is even more apparent when looking at the box plot. As alcohol content goes up, wine quality increases.

```{r echo=FALSE, Bivariate_Plots6}
ggplot(aes(x = quality, y = density), data=ww) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Density (g/cm^3)',
       title = 'Relationship between Density and Quality')
```

This visualization reinforces the negative relationship between density and alcohol. Wines with a higher quality have lower densities.

```{r echo=FALSE, Bivariate_Plots7}
ggplot(aes(x = quality, y = residual.sugar), data=ww) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Residual Sugar (g/dm^3)',
       title = 'Relationship between Residual Sugar and Quality')
```

There is a weak correlation between residual sugar and quality due to low levels of sugar across all wines. Only a few wines have high sugar content.

```{r echo=FALSE, Bivariate_Plots8}
ggplot(aes(x = quality, y = volatile.acidity), data=ww) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Volatile Acidity (g/dm^3)',
       title = 'Relationship between Volatile Acidity and Quality')
```

From the plot, as volatile acidity decreases, quality increases.

```{r echo=FALSE, Bivariate_Plots9}
ggplot(aes(x = quality, y = fixed.acidity), data=ww) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Fixed Acidity (g/dm^3)',
       title = 'Relationship between Fixed Acidity and Quality')
```

As we saw with residual sugar, fixed acidity also has a weak relationship with quality.

```{r echo=FALSE, Bivariate_Plots10}
ggplot(aes(x = quality, y = sulphates), data=ww) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Sulphates (g/dm^3)',
       title = 'Relationship between Sulphates and Quality')
```

It appears that hiigher quality wines have a stronger concentration of sulphates.

```{r echo=FALSE, Bivariate_Plots11}
alcoholQualityLinearModel <- lm(as.numeric(quality) ~ alcohol,
                       data = ww)
summary(alcoholQualityLinearModel)
```

# Bivariate Analysis

### Observations

The plots created in this section support my hypothesis that alcohol plays a role in wine quality. However, my hypothesis that residual sugar affects wine quality was proven wrong. Our findings reveal a strong positive relationship between alcohol and quality but a weak relationship between residual sugar and quality. 

Quality correlates negatively with density and volatile acidity. Wine quality increases as density and volatile acidity decrease. Out of the two variables, density has a stronger correlation with quality.

### Interesting Relationships

I find it interesting that alcohol is negatively correlated with all the variables we plotted (density, residual sugar, volatile acidity).

### Strongest Relationship

Density and residual sugar formed the strongest positive relationship (.8), while density and alcohol had the strongest negative relationship (-.8).

# Multivariate Plots Section

As seen in the previous section, alcohol plays a significant role in wine quality. However alcohol alone can't be the only factor. This leads us to dig deeper to determine how much of an impact alcohol has on quality. 

```{r echo=FALSE, Multivariate_Plots1}
alcoholQualityLinearModel <- lm(as.numeric(quality) ~ alcohol, data = ww)
summary(alcoholQualityLinearModel)
```

According to this model, alcohol contributes to 19% of the total factors affecting quality.
We need to further investigate the relationships among variables to determine other factors.

```{r echo=FALSE, Multivariate_Plots2}
ggplot(data = ww, aes(x = density,  y = alcohol, color = quality)) +
  coord_cartesian(xlim = c(0.985, 1.002), 
                  ylim = c(7.5, 15)) +
  geom_jitter(size=1) +
  geom_smooth(method = 'lm') +
  scale_x_continuous(breaks = seq(0.985, 1.002, 0.002)) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  theme_dark() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Density (g/cm^3)', y = 'Alcohol (% by volume)',
       title = 'Relationship of Density and Alcohol with Quality')
```

Density doesn't appear to significantly change the quality of alcohol. This means the negative correlation between density and quality is due to the presence of alcohol.

```{r echo=FALSE, Multivariate_Plots3}
ggplot(data = ww, aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Alcohol (% by volume)' , y = 'Volatile Acidity (g/dm^3)', 
       title = 'Relationship of Volatile Acidity and Alcohol with Quality')
```

Low concentration of volatile acidity and high alcohol content produce better wines.

```{r echo=FALSE, Multivariate_Plots4}
ggplot(data = ww, aes(x = alcohol, y = sulphates, color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_y_continuous(limits=c(0.3,1.5)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Alcohol (% by volume)' , y = 'Sulphates (g/dm^3)', 
       title = 'Relationship of Sulphates and Alcohol with Quality')
```

Referring back to the correlation matrix, we see that sulphates has the second most positive relationship with quality. After plotting sulphates against alcohol, it appears that higher sulphate concentration and alcohol content produce better wines.

# Multivariate Analysis

### Observations

After looking at the relationships between individual variables with quality, I plotted some variables against alcohol along with quality/rating. This allowed me to determine whether or not these variables have an actual impact on quality. I chose three variables to plot against alcohol: density, volatile acidity, and sulphates.

Density doesn't appear to affect quality. Low volatile acidity and high concentration of sulphates combined with high alcohol content produce better wines. 

### Interesting Relationships

In the Bivariate plot section, we saw that density has a negative relationship with alcohol and quality. Plotting these two variables together allowed me to see the impact of density not just on quality but alcohol as well. This led to an interesting finding: density's correlation with quality is a result of alochol content.

### Model

I created a linear model to calculate alcohol's contribution to quality which is 19%.

------

# Final Plots and Summary

In this section, I chose 3 plots to summarize my findings. 

### Plot 1

```{r echo=FALSE, Plot_1}
ggplot(aes(x = quality, y = alcohol), data = ww) + 
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5, color = 'blue') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Quality', y = 'Alcohol (% by volume)',
       title = 'Relationship between Alcohol and Quality')
```

Alcohol has played a significant role in determining wine quality. The correlation table and this plot reflect the strong positive relationship between alcohol and quality. As alcohol goes up, quality increases.

### Plot 2

```{r echo=FALSE, Plot_2}
ggplot(data = ww, aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Alcohol (% by volume)' , y = 'Volatile Acidity (g/dm^3)', 
       title = 'Relationship of Volatile Acidity and Alcohol with Quality')
```

Volatile acidity may not have as strong of a relationship to quality as alcohol, but it is a factor of quality. A high concentration of alcohol and low concentration of volatile acicity produce higher quality wines.

### Plot 3

```{r echo=FALSE, Plot_3}
ggplot(data = ww, aes(x = alcohol, y = sulphates, color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_y_continuous(limits=c(0.3,1.5)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Alcohol (% by volume)' , y = 'Sulphates (g/dm^3)', 
       title = 'Relationship of Sulphates and Alcohol with Quality')
```

Given the fact that density and residual sugar do not play a major role in quality, I decided to refer back to the correlation table to review correlation of other variables with alcohol and quality. Sulphates stood out to me and the visualization tells us that a higher  concentration of sulphates and alcohol produces better wines.

------

# Reflection

To recap, the objective of this project was to analyze white wine features and their relationships to each other to determine which factors affect our dependent variable, quality.
The approach to investigating the dataset involved using exploratory data analysis. Initially we looked at the variables individually, then we started digging deeper and began looking for insights about relationships among the variables. 

This led to some interesting findings:

• Alcohol has the strongest relationship to quality. As alcohol increases so does quality.
• Variables that make wines taste better include high concentration of alcohol and sulphates and low volatile acidity.

There are limitations to this analysis due to the sampling of only Portuguese wines. The dataset does not include features such as grapes used or wine age. For future analysis, I'd explore a dataset that includes these features. I'd also explore wines from other countries to see if factors affecting quality vary from country to country.

